### 2021-10-03 products/manuscript/Manuscript.Rmd
project proposal in manuscript.rmd and case data

---
title: "Example Manuscript Template for a Data Analysis Project"
subtitle: ""
author: Andreas Handel
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

The structure below is a possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed.

# Summary/Abstract
I plan to investigate the transmission dynamics of SARS-CoV-2 within Houston and between Houston and surrounding areas (the rest of Texas and the United States).

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
As of 10/3/2021, there are 88,776 sequences from Texas uploaded to GISAID (beginning in Feb 2020). After downloading these data, I will align sequences to a reference genome via MAFFT and trim them in Geneious. I will subsample these down to ~400 samples, so that the size is workable in BEAST. Since my focus is on Houston, I will randomly subsample these data while keeping 30% of sequences to Houston.

I will also use case incidence data from the Texas Health Department, as of 10/3/2021.

## Questions/Hypotheses to be addressed
Using case incidence data, assess spatial heterogeneity (county-level) of SARS-CoV-2 in Texas.

Using genomic data, phylogeographic analysis (county-level).

# Methods and Results
Discrete trait analysis (county) in BEAST to estimate transition rates between counties, identify statistically supported transitions (BF values), and calculate R0.

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Exploratory analysis

_Use a combination of text/tables/figures to explore and describe your data. You should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. FIgures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data._

_Continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables. Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones._

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. t-tests, simple regression model with 1 predictor, etc.) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

Table \@ref(tab:summarytable) shows a table summarizing the data.

```{r summarytable,  echo=FALSE}
resulttable=readRDS("../../results/summarytable.rds")
knitr::kable(resulttable, caption = 'Data summary table.')
```

Figure \@ref(fig:resultfigure) shows a scatterplot figure produced by one of the R scripts.

```{r resultfigure,  fig.cap='Analysis figure.', echo=FALSE}
knitr::include_graphics("../../results/resultfigure.png")
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example table \@ref(tab:resulttable) shows a table summarizing a linear model fit.

```{r resulttable,  echo=FALSE}
resulttable=readRDS("../../results/resulttable.rds")
knitr::kable(resulttable, caption = 'Linear model fit table.')
```

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@Leek2015a] discusses types of analyses.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

# References

### 2021-10-08 code/analysis_code/analysisscript.rmd
added data, added to manuscript.rmd, created analysiscode.rmd, generated results (figures and tables)

---
title: "Untitled"
author: "Gabriella Veytsel"
date: "October 8, 2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

processing script
this script loads the raw data, processes and cleans it 
and saves it as Rds file in the processed_data folder

Load packages
```{r}
library(readxl) #for loading Excel files
library(here) #to set paths
library(tidyverse)
library(sf)#st_read to read shapefiles
library(tmap) #plot map
library(scales)
library(broom) #tidy
library(ggpubr) #add equation and r-square 
```

path to data
note the use of the here() package and not absolute paths
```{r}
data_location <- here::here("data","raw_data","CaseCountData.xlsx")

#the first row of data is not column names, skip
rawdata_Cases.and.Fatalities <- read_excel(data_location, sheet = 1, skip = 1) #skip first 2 rows

#the first two rows of data are not column names, skip
rawdata_Trends <- read_excel(data_location, sheet = 2, skip = 3)
```

Take a look at the data
```{r}
#2 obs with missing probable cases, 2 obs with missing confirmed cases
summary(rawdata_Cases.and.Fatalities)

#examine those 2 obs using filter(is.na): 
  #county = unknown & county = grand total
#remove them
cleandata_Cases.and.Fatalities <- rawdata_Cases.and.Fatalities %>% 
  filter(!is.na(`Confirmed Cases`))

summary(rawdata_Trends) #Date should not be a character
head(rawdata_Trends$Date) #Format is year/month/day
cleandata_Trends <- rawdata_Trends %>%
  mutate(Date = as.Date(rawdata_Trends$Date, '%Y/%m/%d'))
head(cleandata_Trends$Date) #looks good
```

Epidemic curve
```{r}
epicurve <- ggplot(cleandata_Trends, aes(x=Date, y = `New Confirmed Cases`)) +
  geom_bar(stat = "identity", fill="lightblue3") +
  scale_x_date(date_breaks = "months" , 
               date_labels = "%b-%y", 
               limits = c(min(cleandata_Trends$Date), max = max(cleandata_Trends$Date)))
epicurve

#Save figure
figure_file_epicurve = here("results","epicurve.png")
ggsave(filename = figure_file_epicurve, plot=epicurve, height = 5, width = 8) 
```

Map
```{r}
#Shapefiles came from: https://www.depts.ttu.edu/geospatial/center/TexasGISData.html
shapefile_location <- here::here("data","raw_data","Tx_CntyBndry_Jurisdictional_TIGER", "Tx_CntyBndry_Jurisdictional_TIGER.shp")
TX_shapefile <- st_read(shapefile_location) #254 features (counties)

#Merge shapefile with metadata
#Concatenate county name with "County" to match shapefile county names
cleandata_Cases.and.Fatalities$County <- paste(cleandata_Cases.and.Fatalities$County, "County", sep = " ")
merge_shp <- merge(TX_shapefile, cleandata_Cases.and.Fatalities,  by.x = "NAME", by.y = "County") 
#the order matters! if shapefile is 2nd, merge_shp is dataframe, not sf
```

There's a population variable (2010 census) in the shapefile! Calcualate population density
```{r}
merge_shp <- merge_shp %>%
  mutate(Population_Density = POP_2010/AREA_SQKM) %>% #population/area (km^2)
  mutate(Population_Density = round(Population_Density,2)) %>%
  arrange(desc(`Confirmed Cases`)) #sort by confirmed cases
```

Examine shapefile
```{r}
ggplot() + geom_sf(data = TX_shapefile)
```

Top 5 counties, most confirmed cases
```{r}
tail(sort(merge_shp$`Confirmed Cases`),5) 
map_cases <- ggplot(merge_shp) +
  geom_sf(aes(fill = `Confirmed Cases`)) +
  ggtitle("Confirmed Cases by County, as of 10/3/2021") +
  geom_sf_text(data = subset(merge_shp, `Confirmed Cases` >= 145433), aes(label = COUNTY), size = 2.5) +
  scale_fill_gradient(low = "light blue", high = "red", name = "Confirmed Cases", labels = comma) 
#labels = comma removes scientific notation (package 'scales')
#label only top 5 counties (in terms of confirmed cases)
#Harris County includes Houston
#Later might want to use arrows

map_cases

#Save figure
figure_file_map_cases = here("results","map_cases.png")
ggsave(filename = figure_file_map_cases, plot=map_cases, height = 12, width = 13) 
```

Top 5 counties, most fatalities
```{r}
tail(sort(merge_shp$Fatalities),5) 
map_deaths <- ggplot(merge_shp) +
  geom_sf(aes(fill = `Fatalities`)) +
  ggtitle("Fatalities by County, as of 10/3/2021") +
  geom_sf_text(data = subset(merge_shp, Fatalities >= 2998), aes(label = COUNTY), size = 2.5) +
  scale_fill_gradient(low = "light blue", high = "red", name = "Fatalities", labels = comma) 
map_deaths

#Save figure
figure_file_map_deaths = here("results","map_deaths.png")
ggsave(filename = figure_file_map_deaths, plot=map_deaths, height = 12, width = 13) 
```

Association between pop density, pop and # confirmed cases, Fatalities? 
```{r}
ggplot(merge_shp, aes(x=POP_2010, y=Fatalities)) +
  geom_point() +
  geom_smooth(method='lm') + #add linear regression line 
  stat_regline_equation(label.y = 8500, aes(label = ..eq.label..)) +
  stat_regline_equation(aes(label = ..rr.label..)) +
  scale_x_continuous(labels = comma) #remove scientific notation

ggplot(merge_shp, aes(x=POP_2010, y=`Confirmed Cases`)) +
  geom_point() +
  geom_smooth(method='lm') + #add linear regression line 
  stat_regline_equation(label.y = 570000, aes(label = ..eq.label..)) +
  stat_regline_equation(aes(label = ..rr.label..)) +
  scale_x_continuous(labels = comma) + #remove scientific notation
  scale_y_continuous(labels = comma)

ggplot(merge_shp, aes(x=Population_Density, y=Fatalities)) +
  geom_point() +
  geom_smooth(method='lm') + #add linear regression line 
  stat_regline_equation(aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 7700, aes(label = ..rr.label..)) +
  scale_x_continuous(labels = comma) #remove scientific notation

ggplot(merge_shp, aes(x=Population_Density, y=`Confirmed Cases`)) +
  geom_point() +
  geom_smooth(method='lm') + #add linear regression line 
  stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
  stat_regline_equation(aes(label = ..rr.label..)) +
  scale_x_continuous(labels = comma) + #remove scientific notation
  scale_y_continuous(labels = comma, limits = c(0, 600000)) 
```

Linear regression
```{r}
lm_deaths_pop <- lm(Fatalities~POP_2010, data = merge_shp)
fit_deaths_pop <- tidy(lm_deaths_pop)
print(fit_deaths_pop)

lm_deaths_density <- lm(Fatalities~Population_Density, data = merge_shp)
fit_deaths_density <- tidy(lm_deaths_density)
print(fit_deaths_density)

lm_cases_pop <- lm(`Confirmed Cases`~POP_2010, data = merge_shp)
fit_cases_pop <- tidy(lm_cases_pop)
print(fit_cases_pop)

lm_cases_density <- lm(`Confirmed Cases`~Population_Density, data = merge_shp)
fit_cases_density <- tidy(lm_cases_density)
print(fit_cases_density)

#Save results tables
table_file_deaths_pop = here("results", "resulttable_deaths_pop.rds")
saveRDS(fit_deaths_pop, file = table_file_deaths_pop)

table_file_deaths_density = here("results", "resulttable_deaths_density.rds")
saveRDS(fit_deaths_density, file = table_file_deaths_density)

table_file_cases_pop = here("results", "resulttable_cases_pop.rds")
saveRDS(fit_cases_pop, file = table_file_cases_pop)

table_file_cases_density = here("results", "resulttable_cases_density.rds")
saveRDS(fit_cases_density, file = table_file_cases_density)
```

save data as RDS
location to save file
```{r}
#save_data_location_spatial <- here::here("data","merge_shp","processeddata_spatial.rds")
#save_data_location_trends <- here::here("data","cleandata_Trends","processeddata_trends.rds")

#saveRDS(merge_shp, file = save_data_location_spatial)
#saveRDS(cleandata_Trends, file = save_data_location_trends)
```

### 2021-10-08 products/manuscript/Manuscript.Rmd
added data, added to manuscript.rmd, created analysiscode.rmd, generated results (figures and tables)

@@
 # Summary/Abstract
-I plan to investigate the transmission dynamics of SARS-CoV-2 within Houston and between Houston and surrounding areas (the rest of Texas and the United States).
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
 
 ## Description of data and data source
-As of 10/3/2021, there are 88,776 sequences from Texas uploaded to GISAID (beginning in Feb 2020).
-After downloading these data, I will align sequences to a reference genome via MAFFT and trim them in Geneious.
-I will subsample these down to ~400 samples, so that the size is workable in BEAST.
-Since my focus is on Houston, I will randomly subsample these data while keeping 30% of sequences to Houston.
+Data from the Texas Health Department, as of 10/3/2021.
+-New confirmed cases and fatalities by date (n = 579)
+-Confirmed cases and fatalities aggregated by county (n = 254)
 
-I will also use case incidence data from the Texas Health Department, as of 10/3/2021.
+Download lineage/clade metadata from GISAID for each county
 
 ## Questions/Hypotheses to be addressed
-Using case incidence data, assess spatial heterogeneity (county-level) of SARS-CoV-2 in Texas.
+Look at spatial heterogeneity within Texas counties (case incidence and mortality)
 
-Using genomic data, phylogeographic analysis (county-level).
+Estimate the reproductive number for each county
+
+Assess the association between population size, population density, and dominant lineage/clade with case incidence, mortality, and reproductive number (regression model)
+
+Use DSAIDE package, regression methods
 
 # Methods and Results
-Discrete trait analysis (county) in BEAST to estimate transition rates between counties, identify statistically supported transitions (BF values), and calculate R0.
 
@@
 
-Figure \@ref(fig:resultfigure) shows a scatterplot figure produced by one of the R scripts.
+Figure \@ref(fig:epicurve) shows an epi curve produced by one of the R scripts.
 
-```{r resultfigure,  fig.cap='Analysis figure.', echo=FALSE}
-knitr::include_graphics("../../results/resultfigure.png")
+```{r epicurve,  fig.cap='Epi Curve.', echo=FALSE}
+knitr::include_graphics("../../results/epicurve.png")
 ```

### 2021-10-08 code/processing_code/processingscript.R

@@ -1,55 +1,158 @@
-###############################
 # processing script
-#
 #this script loads the raw data, processes and cleans it 
 #and saves it as Rds file in the processed_data folder
 
-#load needed packages. make sure they are installed.
+#load packages
 library(readxl) #for loading Excel files
-library(dplyr) #for data processing
 library(here) #to set paths
+library(tidyverse)
+library(sf)#st_read to read shapefiles
+library(tmap) #plot map
+library(scales)
+library(broom) #tidy
+library(ggpubr) #add equation and r-square 
 
 #path to data
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw_data","exampledata.xlsx")
+data_location <- here::here("data","raw_data","CaseCountData.xlsx")
 
-#load data. 
-#note that for functions that come from specific packages (instead of base R)
-# I often specify both package and function like so
-#package::function() that's not required one could just call the function
-#specifying the package makes it clearer where the function "lives",
-#but it adds typing. You can do it either way.
-rawdata <- readxl::read_excel(data_location)
+#the first row of data is not column names, skip
+rawdata_Cases.and.Fatalities <- read_excel(data_location, sheet = 1, skip = 1) #skip first 2 rows
+
+#the first two rows of data are not column names, skip
+rawdata_Trends <- read_excel(data_location, sheet = 2, skip = 3)
 
 #take a look at the data
-dplyr::glimpse(rawdata)
+#2 obs with missing probable cases, 2 obs with missing confirmed cases
+summary(rawdata_Cases.and.Fatalities) 
 
-#dataset is so small, we can print it to the screen.
-#that is often not possible.
-print(rawdata)
+#examine those 2 obs using filter(is.na): 
+  #county = unknown & county = grand total
+#remove them
+cleandata_Cases.and.Fatalities <- rawdata_Cases.and.Fatalities %>% 
+  filter(!is.na(`Confirmed Cases`)) 
 
-# looks like we have measurements for height (in centimeters) and weight (in kilogram)
+summary(rawdata_Trends) #Date should not be a character
+head(rawdata_Trends$Date) #Format is year/month/day
+cleandata_Trends <- rawdata_Trends %>%
[…]

### 2021-10-28 products/manuscript/Manuscript.Rmd
Project Part 3

@@
 ---
-title: "Example Manuscript Template for a Data Analysis Project"
-subtitle: ""
-author: Andreas Handel
-date: "`r Sys.Date()`"
-site: bookdown::bookdown_site
-documentclass: article
+title: 'MADA Data Analysis Project: COVID'
+author: "Andreas Handel"
+date: '`r Sys.Date()`'
 output:
   bookdown::word_document2:
-    toc: false
-    number_sections: true
+    number_sections: yes
+    toc: no
   bookdown::html_document2:
-    toc: false
+    toc: no
+documentclass: article
+csl: ../apa.csl
+site: bookdown::bookdown_site
+subtitle: ''
 bibliography: ../references.bib
-csl: ../apa.csl
 ---
 
-The structure below is a possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed.
-
 # Summary/Abstract
+Understanding the patterns in COVID-19 burden at the county-level can inform targeted public health interventions and resource allocation.
 
 # Introduction 
+Given limited resources, decisions for COVID-19 testing, monitoring, and interventions must be data-driven.
+We aim to characterize the inter-county variability in COVID-19 burden and identify important predictors to help guide public health response and resource allocation, such as vaccine distribution, deployment of mobile testing units, health messaging/education, wastewater monitoring, and hospital equipment.
+Utilizing county-level risk factors in decision-making, in addition to case counts, hospitalizations, and deaths, may allow for a more robust public health response.
 
 ## General Background Information
+Underreporting in cases, due to county-level variation in testing availability and individual-level differences in healthcare access and health-seeking behavior, is a standard consideration in disease reporting.
+With the significant proportion of asymptomatic cases and issues in testing capacity during the early phase, it is a particularly important challenge in the current COVID-19 epidemic.
 
 ## Description of data and data source
-Data from the Texas Health Department, as of 10/3/2021.
--New confirmed cases and fatalities by date (n = 579)
--Confirmed cases and fatalities aggregated by county (n = 254)
+* Data from the Texas Health Department: https://dshs.texas.gov/coronavirus/AdditionalData.aspx
+    + Cases.and.Fatalities: case count data at the county-level, total 
+    + Trends: new cases at the state-level, by date 
+    + NewCasesTime: new cases at the county-level, by date
+    + Demographics: sex, race/ethnicity population
+    + Vaccination: 1-dose and fully vaccinated population
 
-Download lineage/clade metadata from GISAID for each county
+* Shapefile of Texas with county boundaries: https://www.depts.ttu.edu/geospatial/center/TexasGISData.html
+    + Includes data on county population and area to calculate county population density
 
 ## Questions/Hypotheses to be addressed
-Look at spatial heterogeneity within Texas counties (case incidence and mortality)
-
-Estimate the reproductive number for each county
-
-Assess the association between population size, population density, and dominant lineage/clade with case incidence, mortality, and reproductive number (regression model)
-
-Use DSAIDE package, regression methods
+* Investigate the spatial heterogeneity of COVID-19 incidence within Texas over time at the county-level.
+* Evalauate the associations between county-level characteristics (e.g., population density, population demographics, and vaccination coverage) and COVID-19 burden.
 
 # Methods and Results
-
-## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+* Estimate county-level cumulative rates of COVID-19: new cases per 100,000 county residents
+* Test for spatial autocorrelation using Moran's I
+* Multivariable regression, with spatial effects, using negative binomial regression to account for overdispersion (excess zeroes)
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-You should produce plots or tables or other summary quantities for the most interesting/important quantities in your data.
-Depending on the total number of variables in your dataset, explore all or some of the others.
-FIgures produced here might be histograms or density plots, correlation plots, etc.
-Tables might summarize your data._
+Cases over time for Texas.
 
-_Continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in.
-If your dataset is small, you can do that for all variables.
-Plots produced here can be scatterplots, boxplots, violinplots, etc.
-Tables can be simple 2x2 tables or larger ones._
-
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-t-tests, simple regression model with 1 predictor, etc.) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
-
-Table \@ref(tab:summarytable) shows a table summarizing the data.
-
-```{r summarytable,  echo=FALSE}
-resulttable=readRDS("../../results/summarytable.rds")
-knitr::kable(resulttable, caption = 'Data summary table.')
-```
-
-Figure \@ref(fig:epicurve) shows an epi curve produced by one of the R scripts.
-
-```{r epicurve,  fig.cap='Epi Curve.', echo=FALSE}
+```{r epicurve,  fig.cap='', echo=FALSE}
 knitr::include_graphics("../../results/epicurve.png")
@@
 
+County-level characteristics across Texas
+
+```{r summarytable,  results = 'asis', echo=FALSE, message = FALSE, warning = FALSE}
+library(tidyverse)
+
+resulttable=readRDS("../../results/summarytable.rds")
+knitr::kable(resulttable, caption = '') %>%
+  kableExtra::kable_classic(full_width = F, html_font = "Times New Roman")
+```
+
+County-level characteristics for Texas counties with the highest COVID-19 case burden
+
+```{r table_county,  results = 'asis', echo=FALSE, echo=FALSE, message = FALSE, warning = FALSE}
+library(tidyverse)
+
+resulttable=readRDS("../../results/table_county.rds")
+knitr::kable(resulttable, caption = '') %>%
+  kableExtra::kable_classic(full_width = F, html_font = "Times New Roman")
+```
+
+Map of Texas, highlighting counties with the highest COVID-19 case burden
+
+```{r map_rates,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/map_rates.png")
+```
+
+Cases over time for Texas counties with the highest COVID-19 case burden
+
+```{r epicurve_county,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/epicurve_county.png")
+```
+
+Association between county-level characteristics and total confirmed COVID-19 cases.
+
+```{r associations,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/p_combined.png")
+```
+
 ## Full analysis
+Output of a mixed effects model
 
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
-Example table \@ref(tab:resulttable) shows a table summarizing a linear model fit.
-
-```{r resulttable,  echo=FALSE}
-resulttable=readRDS("../../results/resulttable.rds")
-knitr::kable(resulttable, caption = 'Linear model fit table.')
+```{r glm_mixed,  fig.cap='GLM Random Effects Model.', echo=FALSE, message = FALSE, warning = FALSE}
+library(sjPlot)
+model=readRDS("../../results/model.rds")
+tab_model(model) 
 ```
@@
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+There are many potentially important predictors that were not considered in this study, such as county-level proportions for urban/rural, medical conditions, household size,  socio-economic status, and frontline workers.
+While this study model used cumulative case counts, future studies investigating county-level case counts over time should also examine variant proportions and testing rates as predictors.
 

### 2021-10-28 code/analysis_code/analysisscript.R

@@ -1,0 +1,56 @@
+#Outcome: county-level cumulative rates of COVID-19
+#County-level predictors of interest: 
+  #Population Size
+  #Population Density
+  #Population, 65+
+  #Population, Vaccinated (1-dose and full)
+  #Population, Black
+  #Population, Hispanic
+  #Population, Asian
+  #Population, Male
+
+#I don't have vaccination data by county over time, only vaccination x county & vaccination x time
+#I don't have good variant data by county over time, only variant over time in Texas
+  #Future analysis: outocme = weekly county-level incidence data, predictor = weekly variant proportion
+
+library(tidymodels)
+library(poissonreg)
+library(lme4)
+library(sjPlot)
+
+#I don't see an offset option in tidymodels
+#??????????????????????????????????????????
+
+#Does it make more sense to use counts and offset
+#Or convert daily incidence to monthly incidence
+#???????????????????????????????????????????????
+
+#Generalized linear mixed model
+#Family = poisson (count data)
+#Random effects = county: assumption that observations within counties are correlated
+#Is adding random effects necessary when each observation is already a county? 
+#Still doesn't account for correlation *between* counties: may need spatial component
+#Offset = population size
+#Scale: continuous covariates were scaled to have mean 0 and SD 1
+model <- glmer(formula = `Confirmed Cases` ~ offset(log(Population)) + #offset = population size
+              scale(Percent65Plus) + 
+              scale(PercentHispanic) +
+              scale(PercentBlack) +
+              scale(PercentAsian) +
+              scale(PercentVaccinated_1dose) +
+              scale(PercentVaccinated_full) +
+              scale(Population_Density) +
+              (1|COUNTY), #random effects = county
+              family=poisson, #count data
+              data = merge_shp_vac_df_c) 
+summary(model)
+fixef(model)
+model_results <- tab_model(model)
+
+#Save table
+save_data_location1 <- here::here("results","model.rds")
+saveRDS(model, file = save_data_location1)
+
+#Investigate spatial autocorrelation: calculate Moran's I
+#********************************************************
+

### 2021-10-28 code/analysis_code/explorationscript.R

@@ -1,0 +1,217 @@
+#load packages
+library(readxl) #for loading Excel files
+library(here) #to set paths
+library(tidyverse)
+library(sf)#st_read to read shapefiles
+library(tmap) #plot map
+library(scales)
+library(broom) #tidy
+library(ggpubr) #add equation and r-square 
+library(tidyr)
+library(cowplot)
+library(table1) #html table 1
+
+#Read RDS files from processingscript.R
+#**************************************
+
+save_data_location1 <- here::here("data","processed_data", "merge_shp_vac.rds")
+save_data_location2 <- here::here("data","processed_data","cleandata_Trends.rds")
+save_data_location3 <- here::here("data","processed_data","cleandata_Cases.and.Fatalities.rds")
+save_data_location4 <- here::here("data","processed_data","cleandata_Vaccination.rds")
+save_data_location5 <- here::here("data","processed_data","NewCasesTime_long.rds")
+save_data_location6 <- here::here("data","processed_data","cleandata_Demographics.rds")
+
+merge_shp_vac <- readRDS(save_data_location1)
+cleandata_Trends <- readRDS(save_data_location2)
+cleandata_Cases.and.Fatalities <- readRDS(save_data_location3)
+cleandata_Vaccination <- readRDS(save_data_location4)
+NewCasesTime_long <- readRDS(save_data_location5)
+cleandata_Demographics <- readRDS(save_data_location6)
+
+#Epidemic curve
+#**************
+
+epicurve <- ggplot(cleandata_Trends, aes(x=Date, y = `New Confirmed Cases`)) +
+  geom_bar(stat = "identity", fill="lightblue3") +
+  scale_x_date(date_breaks = "months" , 
+               date_labels = "%b-%y", 
+               limits = c(min(cleandata_Trends$Date), max = max(cleandata_Trends$Date)))
+
+epicurve
+
+#Save figure
+figure_file_epicurve = here("results","epicurve.png")
+ggsave(filename = figure_file_epicurve, plot=epicurve, height = 5, width = 8) 
+
+#Epidemic Curve, by County
+#*************************
+
+#Too many counties for facet_wrap, narrow down to top 10 counties
+NewCasesTime_long %>%
+  group_by(County) %>%
+  summarize(Sum = sum(NewCases)) %>%
[…]

### 2021-10-28 code/processing_code/processingscript.R

@@ -11,26 +11,49 @@ library(tmap) #plot map
 library(scales)
 library(broom) #tidy
 library(ggpubr) #add equation and r-square 
+library(tidyr)
+library(cowplot)
+library(table1) #html table 1
 
 #path to data
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw_data","CaseCountData.xlsx")
+data_location1 <- here::here("data","raw_data","CaseCountData.xlsx")
+data_location2 <- here::here("data","raw_data","COVID-19 Vaccine Data by County.xlsx")
+data_location3 <- here::here("data","raw_data","Texas COVID-19 New Confirmed Cases by County.xlsx")
+data_location4 <- here::here("data","raw_data","Population Demographics.csv")
+
+#Population Demographics, Texas
+#******************************
+rawdata_Demographics <- read_csv(data_location4)
+cleandata_Demographics <- rawdata_Demographics %>%
+  filter(Age == "All Ages") %>% 
+  filter(FIPS != "000")
+
+glimpse(cleandata_Demographics)
+
+#Total Cases and Fatalities
+#**************************
 
 #the first row of data is not column names, skip
-rawdata_Cases.and.Fatalities <- read_excel(data_location, sheet = 1, skip = 1) #skip first 2 rows
-
-#the first two rows of data are not column names, skip
-rawdata_Trends <- read_excel(data_location, sheet = 2, skip = 3)
+rawdata_Cases.and.Fatalities <- read_excel(data_location1, sheet = 1, skip = 1) #skip first 2 rows
 
 #take a look at the data
 #2 obs with missing probable cases, 2 obs with missing confirmed cases
 summary(rawdata_Cases.and.Fatalities) 
 
 #examine those 2 obs using filter(is.na): 
-  #county = unknown & county = grand total
+#county = unknown & county = grand total
 #remove them
 cleandata_Cases.and.Fatalities <- rawdata_Cases.and.Fatalities %>% 
   filter(!is.na(`Confirmed Cases`)) 
+
+glimpse(cleandata_Cases.and.Fatalities)
+
+#New Confirmed Cases over Time
+#*****************************
+
+#the first two rows of data are not column names, skip
+rawdata_Trends <- read_excel(data_location1, sheet = 2, skip = 3)
 
[…]

### 2021-11-15 code/analysis_code/explorationscript.Rmd
Project Part 4

---
title: "explorationscript"
author: "Gabriella Veytsel"
date: "November 12, 2021"
output: html_document
---
#load packages
```{r}
library(readxl) #for loading Excel files
library(here) #to set paths
library(tidyverse)
library(sf)#st_read to read shapefiles
library(tmap) #plot map
library(scales)
library(broom) #tidy
library(ggpubr) #add equation and r-square 
library(tidyr)
library(cowplot)
library(table1) #html table 1
```

#Read RDS files from processingscript.R
```{r}
save_data_location1 <- here::here("data","processed_data","merge_shp_vac.rds")
save_data_location2 <- here::here("data","processed_data","cleandata_Trends.rds")
save_data_location3 <- here::here("data","processed_data","cleandata_Cases.and.Fatalities.rds")
save_data_location4 <- here::here("data","processed_data","cleandata_Vaccination.rds")
save_data_location5 <- here::here("data","processed_data","NewCasesTime_long.rds")
save_data_location6 <- here::here("data","processed_data","cleandata_Demographics.rds")

merge_shp_vac <- readRDS(save_data_location1)
cleandata_Trends <- readRDS(save_data_location2)
cleandata_Cases.and.Fatalities <- readRDS(save_data_location3)
cleandata_Vaccination <- readRDS(save_data_location4)
NewCasesTime_long <- readRDS(save_data_location5)
cleandata_Demographics <- readRDS(save_data_location6)
```

#Epidemic curve
```{r}
epicurve <- ggplot(cleandata_Trends, aes(x=Date, y = `New Confirmed Cases`)) +
  geom_bar(stat = "identity", fill="lightblue3") +
  scale_x_date(date_breaks = "months" , 
               date_labels = "%b-%y", 
               limits = c(min(cleandata_Trends$Date), max = max(cleandata_Trends$Date)))

epicurve

#Save figure
figure_file_epicurve = here("results","epicurve.png")
ggsave(filename = figure_file_epicurve, plot=epicurve, height = 5, width = 8) 
```

#Epidemic Curve, by County
```{r}
#Too many counties for facet_wrap, narrow down to top 10 counties
NewCasesTime_long %>%
  group_by(County) %>%
  summarize(Sum = sum(NewCases)) %>%
  arrange(desc(Sum))

NewCasesTime_long_subset <- NewCasesTime_long %>%
  filter(County %in% c("Harris", "Dallas", "Tarrant", "Bexar", "El Paso", "Travis", "Collin", "Fort Bend", "Hidalgo", "Denton"))

epicurve_county <- ggplot(NewCasesTime_long_subset, aes(x=Date, y = NewCases)) +
  geom_bar(stat = "identity", fill="lightblue3") +
  scale_x_date(date_breaks = "months" , 
               date_labels = "%b-%y", 
               limits = c(min(NewCasesTime_long$Date), max = max(NewCasesTime_long$Date))) + 
  facet_wrap(~County, nrow = 2) + #2 rows
  theme(axis.text.x = element_text(angle=50, hjust=1)) +
  labs(title = "Epidemic Curve", 
       subtitle = "Subset to counties with the most total cases in Texas",
       y = "New Cases", x = "")

epicurve_county

#Save figure
figure_file_epicurve_county = here("results","epicurve_county.png")
ggsave(filename = figure_file_epicurve_county, plot=epicurve_county, height = 5, width = 15) 
```

#Confirmed Case, by County
```{r}
tail(sort(merge_shp_vac$`Confirmed Cases`),10) #top 10 counties with the most confirmed cases
map_cases <- ggplot(merge_shp_vac) +
  geom_sf(aes(fill = `Confirmed Cases`)) +
  ggtitle("Confirmed Cases by County, as of 10/3/2021") +
  geom_sf_text(data = subset(merge_shp_vac, `Confirmed Cases` >= 67896), aes(label = COUNTY), size = 2.5) +
  scale_fill_gradient(low = "light blue", high = "red", name = "Confirmed Cases", labels = comma) 
#labels = comma removes scientific notation (package 'scales')

map_cases
#As expected, these are all the major cities. Instead, look at rates to compare counties with different population sizes

#Save figure
figure_file_map_cases = here("results","map_cases.png")
ggsave(filename = figure_file_map_cases, plot=map_cases, height = 12, width = 15) 
```

#Cumulative Confirmed Case Rate/100 Persons, by County
```{r}
tail(sort(merge_shp_vac$CumCaseRatePer100),10) #top 10 counties with the highest cumulative case rate
map_rates <- ggplot(merge_shp_vac) +
  geom_sf(aes(fill = CumCaseRatePer100)) +
  ggtitle("Cumulative Confirmed Case Rate/100 Persons, by County, as of 10/3/2021") +
  geom_sf_text(data = subset(merge_shp_vac, CumCaseRatePer100 >= 191), aes(label = COUNTY), size = 2.5) +
  scale_fill_gradient(low = "light blue", high = "red", name = "Cumulative Confirmed Case Rate/100 Persons", labels = comma)

map_rates
#Very different results, hmm...let's summarize county-level characteristics

#Save figure
figure_file_map_rates = here("results","map_rates.png")
ggsave(filename = figure_file_map_rates, plot=map_rates, height = 12, width = 15) 
```

```{r}
merge_shp_vac %>%
  dplyr::select(COUNTY, CumCaseRatePer100) %>%
  arrange(desc(CumCaseRatePer100))

merge_shp_vac_subset <- merge_shp_vac %>%
  dplyr::select(COUNTY, `Confirmed Cases`, Total, CumCaseRatePer100, Population_Density, Percent65Plus, PercentVaccinated_1dose, PercentVaccinated_full, PercentHispanic, PercentAsian, PercentBlack, PercentMale, AREA_SQKM, PercentMedCondition) %>%
  rename(Population = Total)
summary(merge_shp_vac_subset)

merge_shp_vac_df <- as.data.frame(merge_shp_vac_subset) %>%
  dplyr::select(-c(COUNTY, geometry))

merge_shp_vac_df_c <- as.data.frame(merge_shp_vac_subset) %>%
  dplyr::select(-c(geometry)) #has to be a dataframe without geometry for table 1

#Save dataset
save_data_location10 <- here::here("data","processed_data","merge_shp_vac_df_c.rds")
saveRDS(merge_shp_vac_df_c, file = save_data_location10)

table1 <- table1(~ . , data = merge_shp_vac_df, topclass="Rtable1-zebra") 
table1

#Save table
save_data_location7 <- here::here("results","summarytable.rds")
saveRDS(table1, file = save_data_location7)
```

#County characteristics, subset to counties with highest cumulative case rates
```{r}
table_county <- merge_shp_vac_df_c %>%
  filter(COUNTY %in% c("Dimmit", "Hale", "Uvalde", "Maverick", "Karnes", "Childress", "Lubbock", "Val Verde", "Potter", "Chambers" )) %>%
  group_by(COUNTY) %>%
  arrange(desc(CumCaseRatePer100))

#Save table
save_data_location6 <- here::here("results","table_county.rds")
saveRDS(table_county, file = save_data_location6)
```

#Association between county-level characteristics and # confirmed cases 
```{r}
p1 <- ggscatter(merge_shp_vac_subset, x="PercentHispanic", y="CumCaseRatePer100",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "spearman")
p2 <- ggscatter(merge_shp_vac_subset, x="Percent65Plus", y="CumCaseRatePer100",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "spearman")
p3 <- ggscatter(merge_shp_vac_subset, x="PercentMale", y="CumCaseRatePer100",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "spearman")
p4 <- ggscatter(merge_shp_vac_subset, x="PercentBlack", y="CumCaseRatePer100",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "spearman")
p5 <- ggscatter(merge_shp_vac_subset, x="PercentAsian", y="CumCaseRatePer100",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "spearman")
p6 <- ggscatter(merge_shp_vac_subset, x="PercentMedCondition", y="CumCaseRatePer100",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "spearman")
p7 <- ggscatter(merge_shp_vac_subset, x="AREA_SQKM", y="CumCaseRatePer100",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "spearman")

p_combined <- plot_grid(p1, p2, p3, p4, p5, p6, p7)

# Save figure
figure_file_association = here("results","p_combined.png")
ggsave(filename = figure_file_association, plot=p_combined, height = 5, width = 15.5)

```

```{r}
# p1 <- ggplot(merge_shp_vac_subset, aes(x=Population, y=`Confirmed Cases`)) +
#   geom_point() +
#   geom_smooth(method='lm') + #add linear regression line 
#   #stat_regline_equation(label.y = 570000, aes(label = ..eq.label..)) +
#   stat_regline_equation(aes(label = ..rr.label..)) +
#   scale_x_continuous(labels = comma) + #remove scientific notation
#   scale_y_continuous(labels = comma) 
# 
# p2 <- ggplot(merge_shp_vac_subset, aes(x=PercentMale, y=CumCaseRatePer100)) +
#   geom_point() +
#   geom_smooth(method='lm') + #add linear regression line 
#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
#   stat_regline_equation(aes(label = ..rr.label..)) +
#   scale_x_continuous(labels = comma) + #remove scientific notation
#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
# 
# p3 <- ggplot(merge_shp_vac_subset, aes(x=PercentBlack, y=CumCaseRatePer100)) +
#   geom_point() +
#   geom_smooth(method='lm') + #add linear regression line 
#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
#   stat_regline_equation(aes(label = ..rr.label..)) +
#   scale_x_continuous(labels = comma) + #remove scientific notation
#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
# 
# p4 <- ggplot(merge_shp_vac_subset, aes(x=PercentAsian, y=CumCaseRatePer100)) +
#   geom_point() +
#   geom_smooth(method='lm') + #add linear regression line 
#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
#   stat_regline_equation(aes(label = ..rr.label..)) +
#   scale_x_continuous(labels = comma) + #remove scientific notation
#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
# 
# p5 <- ggplot(merge_shp_vac, aes(x=PercentHispanic, y=CumCaseRatePer100)) +
#   geom_point() +
#   geom_smooth(method='lm') + #add linear regression line 
#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
#   stat_regline_equation(aes(label = ..rr.label..)) +
#   scale_x_continuous(labels = comma) + #remove scientific notation
#   scale_y_continuous(labels = comma, limits = c(0,30))
# 
# p6 <- ggplot(merge_shp_vac_subset, aes(x=Population_Density, y=CumCaseRatePer100)) +
#   geom_point() +
#   geom_smooth(method='lm') + #add linear regression line 
#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
#   stat_regline_equation(aes(label = ..rr.label..)) +
#   scale_x_continuous(labels = comma) + #remove scientific notation
#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
# 
# p7 <- ggplot(merge_shp_vac_subset, aes(x=Percent65Plus, y=CumCaseRatePer100)) +
#   geom_point() +
#   geom_smooth(method='lm') + #add linear regression line 
#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
#   stat_regline_equation(aes(label = ..rr.label..)) +
#   scale_x_continuous(labels = comma) + #remove scientific notation
#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
# 
# p8 <- ggplot(merge_shp_vac_subset, aes(x=PercentVaccinated_1dose, y=CumCaseRatePer100)) +
#   geom_point() +
#   geom_smooth(method='lm') + #add linear regression line 
#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
#   stat_regline_equation(aes(label = ..rr.label..)) +
#   scale_x_continuous(labels = comma) + #remove scientific notation
#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
# 
# p9 <- ggplot(merge_shp_vac_subset, aes(x=PercentVaccinated_full, y=CumCaseRatePer100)) +
#   geom_point() +
#   geom_smooth(method='lm') + #add linear regression line 
#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
#   stat_regline_equation(aes(label = ..rr.label..)) +
#   scale_x_continuous(labels = comma) + #remove scientific notation
#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
# 
# p_combined <- plot_grid(p1, p2, p3, p4, p5, p6, p7, p8, p9) 
# 
# #Save figure
# figure_file_association = here("results","p_combined.png")
# ggsave(filename = figure_file_association, plot=p_combined, height = 5, width = 15.5)

```

### 2021-11-15 products/manuscript/Manuscript.Rmd
Project Part 4

@@
 output:
+  bookdown::html_document2:
+    toc: no
   bookdown::word_document2:
     number_sections: yes
-    toc: no
-  bookdown::html_document2:
     toc: no
@@
     + NewCasesTime: new cases at the county-level, by date
-    + Demographics: sex, race/ethnicity population
-    + Vaccination: 1-dose and fully vaccinated population
+    + Demographics: sex, race/ethnicity
+    + Vaccination: 1-dose and fully vaccinated population, population 65+ from 2019 Census population estimates, population 16-64 any medical condition*
 
@@
 # Methods and Results
-* Estimate county-level cumulative rates of COVID-19: new cases per 100,000 county residents
-* Test for spatial autocorrelation using Moran's I
-* Multivariable regression, with spatial effects, using negative binomial regression to account for overdispersion (excess zeroes)
+Cumulative numbers of COVID-19 cases and 2019 population estimates for each county were obtained from the Texas Department of Health.
+As counties containing major cities (e.g., Houston, Dallas, Fort Worth) had the highest total cases, cumulative case rates per 100 persons were calculated and mapped to visualize spatial heterogeneity across Texas counties.
+Epidemic curves were constructed to visualize new cases over time at the state- and county- levels (subset to 10 counties with the highest total case burden).
+County-level risk factors included in our analysis were: age (65 and older), gender (male), race/ethnicity (Black, Asian, and Hispanic), and any medical condition that increases risk of COVID*.
+
+To estimate the association between county-level risk factors and cumulative case incidence, the following models were developed: Poisson regression, negative binomial regression, multivariable linear regression, and a generalized mixed effects model.
+For the Poisson model, the assumption that the mean and variance of number of cumulative COVID-19 cases were approximately equal was violated; additionally, there was evidence of severe overdispersion.
+Ultimately, this model was rejected as an appropriate model choice.
+The negative binomial regression model was included to relax the assumption of mean-variance equivalence and, accordingly, had superior performance.
+Several diagnostic plots (fitted vs.
+residual plots, normal QQ plot, scale-location, and residuals vs.
+leverage) were inspected to test if linear regression model assumptions were met.
+Indeed, the assumptions of linearity of the predictor and outcome, constant variance of residuals, normality of residuals, and independence of residuals error terms of the multivariable linear regression were accepted.
+
+We used the TidyModels framework in R to build a penalized linear regression model (LASSO) and Random Forest regression model.
+LASSO removed the proportion male variable and applied a penality of 0.1 to remaining predictors.
+While the Random Forest model does not remove predictors, the VIP package in R also selected PercentMale as the least important variable.
+The best model for each approach was chosen based on RMSE and diagnostic plots, which was very similar (LASSO RMSE = 3.82, RF RMSE = 3.76).
+We selected the LASSO model as the final model and conducted 5-fold cross-validation on the test dataset one single time.
+
+*People 16 years of age and older with at least one chronic medical condition that puts them at increased risk for severe illness from the virus that causes COVID-19.
+The data are model-based estimates from CDC using self-reported data from the Behavioral Risk Factor Surveillance System and U.S.
+census population data.
+The methodology for generating these estimates are detailed in this MMWR: https://www.cdc.gov/mmwr/volumes/69/wr/mm6929a1.htm.
 
@@
 
+Map of Texas, highlighting counties with the highest COVID-19 case burden
+
+```{r map_rates,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/map_rates.png")
+```
+
 County-level characteristics for Texas counties with the highest COVID-19 case burden
@@
 
-Map of Texas, highlighting counties with the highest COVID-19 case burden
-
-```{r map_rates,  fig.cap='', echo=FALSE}
-knitr::include_graphics("../../results/map_rates.png")
-```
-
 Cases over time for Texas counties with the highest COVID-19 case burden
@@
 Association between county-level characteristics and total confirmed COVID-19 cases.
+Spearman's correlation coefficient and regression line
 
@@
 ## Full analysis
-Output of a mixed effects model
+Negative Binomial Model Evaluation
 
-```{r glm_mixed,  fig.cap='GLM Random Effects Model.', echo=FALSE, message = FALSE, warning = FALSE}
-library(sjPlot)
-model=readRDS("../../results/model.rds")
-tab_model(model) 
+```{r NegBinomial_diagnosticplots,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/NegBinomial_diagnosticplots.jpeg")
+```
+
+```{r NegBinomial_pred_vs_obs,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/NegBinomial_pred_vs_obs.jpeg")
+```
+
+```{r NegBinomialresiduals,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/NegBinomial_residuals.jpeg")
+```
+
+Generalized Linear Mixed Model Evaluation
+
+```{r mixed_pred_vs_obs,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/MixedModel_pred_vs_obs.jpeg")
+```
+
+```{r mixedresiduals,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/MixedModel_residuals.jpeg")
+```
+
+LASSO Model Evaluation
+
+```{r LASSO_pred_vs_obs,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/LASSO_predicted_vs_observed.jpeg")
+```
+
+```{r LASSOresiduals,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/LASSO_residuals.jpeg")
+```
+
+Random Forest Model Evaluation
+
+```{r rf_pred_vs_obs,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/RF_pred_vs_obs.jpeg")
+```
+
+```{r rfresiduals,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/RF_residuals.jpeg")
 ```
@@
 ## Strengths and Limitations
-There are many potentially important predictors that were not considered in this study, such as county-level proportions for urban/rural, medical conditions, household size,  socio-economic status, and frontline workers.
-While this study model used cumulative case counts, future studies investigating county-level case counts over time should also examine variant proportions and testing rates as predictors.
+There are many potentially important predictors that were not considered in this study, such as county-level proportions for urban/rural, household size, and socio-economic status.
 

### 2021-11-15 code/analysis_code/analysisscript.R

@@ -1,56 +1,350 @@
 #Outcome: county-level cumulative rates of COVID-19
 #County-level predictors of interest: 
-  #Population Size
-  #Population Density
-  #Population, 65+
-  #Population, Vaccinated (1-dose and full)
-  #Population, Black
-  #Population, Hispanic
-  #Population, Asian
-  #Population, Male
+#Population Size
+#Population Density
+#Population, 65+
+#Population, Black
+#Population, Hispanic
+#Population, Asian
+#Population, Male
+#Population, Medical Condition
 
-#I don't have vaccination data by county over time, only vaccination x county & vaccination x time
-#I don't have good variant data by county over time, only variant over time in Texas
-  #Future analysis: outocme = weekly county-level incidence data, predictor = weekly variant proportion
-
+#Load packages
+library(tidyverse)
 library(tidymodels)
 library(poissonreg)
 library(lme4)
+library(broom)
 library(sjPlot)
+library(ggpubr)
+library(performance) #overdispersion
+library(AER) #overdispersion
+library(MASS) #negative binomial
+library(vip)
 
-#I don't see an offset option in tidymodels
-#??????????????????????????????????????????
+#Load data
+save_data_location1 <- here::here("data","processed_data","merge_shp_vac_df_c.rds")
+save_data_location2 <- here::here("data","processed_data","cleandata_Trends.rds")
+save_data_location3 <- here::here("data","processed_data","cleandata_Cases.and.Fatalities.rds")
+save_data_location4 <- here::here("data","processed_data","cleandata_Vaccination.rds")
+save_data_location5 <- here::here("data","processed_data","NewCasesTime_long.rds")
+save_data_location6 <- here::here("data","processed_data","cleandata_Demographics.rds")
 
-#Does it make more sense to use counts and offset
-#Or convert daily incidence to monthly incidence
-#???????????????????????????????????????????????
+combined <- readRDS(save_data_location1)
+cleandata_Trends <- readRDS(save_data_location2)
+cleandata_Cases.and.Fatalities <- readRDS(save_data_location3)
+cleandata_Vaccination <- readRDS(save_data_location4)
+NewCasesTime_long <- readRDS(save_data_location5)
+cleandata_Demographics <- readRDS(save_data_location6)
 
-#Generalized linear mixed model
-#Family = poisson (count data)
[…]

### 2021-11-15 code/analysis_code/explorationscript.R

@@ -1,217 +1,0 @@
-#load packages
-library(readxl) #for loading Excel files
-library(here) #to set paths
-library(tidyverse)
-library(sf)#st_read to read shapefiles
-library(tmap) #plot map
-library(scales)
-library(broom) #tidy
-library(ggpubr) #add equation and r-square 
-library(tidyr)
-library(cowplot)
-library(table1) #html table 1
-
-#Read RDS files from processingscript.R
-#**************************************
-
-save_data_location1 <- here::here("data","processed_data", "merge_shp_vac.rds")
-save_data_location2 <- here::here("data","processed_data","cleandata_Trends.rds")
-save_data_location3 <- here::here("data","processed_data","cleandata_Cases.and.Fatalities.rds")
-save_data_location4 <- here::here("data","processed_data","cleandata_Vaccination.rds")
-save_data_location5 <- here::here("data","processed_data","NewCasesTime_long.rds")
-save_data_location6 <- here::here("data","processed_data","cleandata_Demographics.rds")
-
-merge_shp_vac <- readRDS(save_data_location1)
-cleandata_Trends <- readRDS(save_data_location2)
-cleandata_Cases.and.Fatalities <- readRDS(save_data_location3)
-cleandata_Vaccination <- readRDS(save_data_location4)
-NewCasesTime_long <- readRDS(save_data_location5)
-cleandata_Demographics <- readRDS(save_data_location6)
-
-#Epidemic curve
-#**************
-
-epicurve <- ggplot(cleandata_Trends, aes(x=Date, y = `New Confirmed Cases`)) +
-  geom_bar(stat = "identity", fill="lightblue3") +
-  scale_x_date(date_breaks = "months" , 
-               date_labels = "%b-%y", 
-               limits = c(min(cleandata_Trends$Date), max = max(cleandata_Trends$Date)))
-
-epicurve
-
-#Save figure
-figure_file_epicurve = here("results","epicurve.png")
-ggsave(filename = figure_file_epicurve, plot=epicurve, height = 5, width = 8) 
-
-#Epidemic Curve, by County
-#*************************
-
-#Too many counties for facet_wrap, narrow down to top 10 counties
-NewCasesTime_long %>%
-  group_by(County) %>%
-  summarize(Sum = sum(NewCases)) %>%
[…]

### 2021-11-15 code/processing_code/processingscript.R

@@ -120,7 +120,7 @@ merge_shp <- merge_shp %>%
 
 #Add cumulative case rates
 merge_shp <- merge_shp %>%
-  mutate(CumCaseRatePer1000 = `Confirmed Cases`/Total * 1000)
+  mutate(CumCaseRatePer100 = `Confirmed Cases`/Total * 100)
 
 #Add vaccination
 #Calculate proportions
@@ -132,9 +132,14 @@ merge_shp_vac <- merge_shp %>%
   mutate(PopulationVaccinated_full = as.numeric(`People Fully Vaccinated`)) %>%
   mutate(PercentVaccinated_full = PopulationVaccinated_full /  Total * 100) %>%
   
+  rename(PopulationMedCondition = `Population, 16-64\r\n Any Medical Condition`) %>%
+  mutate(PopulationMedCondition = as.numeric(PopulationMedCondition)) %>%
+  mutate(PercentMedCondition = PopulationMedCondition /  Total * 100) %>%
+    
   mutate(PercentHispanic = Hispanic_Total /  Total * 100) %>%
   mutate(PercentAsian = NH_Asian_Total /  Total * 100) %>%
-  mutate(PercentBlack = NH_Black_Total /  Total * 100)
+  mutate(PercentBlack = NH_Black_Total /  Total * 100) %>%
+  mutate(PercentMale = Total_Male / Total * 100)
 
 # save data as RDS
 # location to save file

### 2021-11-29 code/analysis_code/explorationscript.Rmd
Project part 5

@@
 ---
-#load packages
+Load packages
 ```{r}
@@
 
-#Read RDS files from processingscript.R
+Read RDS files from processingscript.R
 ```{r}
@@
 
-#Epidemic curve
+Epidemic curve
 ```{r}
@@
                date_labels = "%b-%y", 
-               limits = c(min(cleandata_Trends$Date), max = max(cleandata_Trends$Date)))
+               limits = c(min(cleandata_Trends$Date), max = max(cleandata_Trends$Date))) + 
+  theme(axis.text.x = element_text(angle = 45))
 
@@
 
-#Epidemic Curve, by County
-```{r}
-#Too many counties for facet_wrap, narrow down to top 10 counties
-NewCasesTime_long %>%
-  group_by(County) %>%
-  summarize(Sum = sum(NewCases)) %>%
-  arrange(desc(Sum))
-
-NewCasesTime_long_subset <- NewCasesTime_long %>%
-  filter(County %in% c("Harris", "Dallas", "Tarrant", "Bexar", "El Paso", "Travis", "Collin", "Fort Bend", "Hidalgo", "Denton"))
-
-epicurve_county <- ggplot(NewCasesTime_long_subset, aes(x=Date, y = NewCases)) +
-  geom_bar(stat = "identity", fill="lightblue3") +
-  scale_x_date(date_breaks = "months" , 
-               date_labels = "%b-%y", 
-               limits = c(min(NewCasesTime_long$Date), max = max(NewCasesTime_long$Date))) + 
-  facet_wrap(~County, nrow = 2) + #2 rows
-  theme(axis.text.x = element_text(angle=50, hjust=1)) +
-  labs(title = "Epidemic Curve", 
-       subtitle = "Subset to counties with the most total cases in Texas",
-       y = "New Cases", x = "")
-
-epicurve_county
-
-#Save figure
-figure_file_epicurve_county = here("results","epicurve_county.png")
-ggsave(filename = figure_file_epicurve_county, plot=epicurve_county, height = 5, width = 15) 
-```
-
-#Confirmed Case, by County
+Confirmed Case, by County
 ```{r}
@@
 tail(sort(merge_shp_vac$CumCaseRatePer100),10) #top 10 counties with the highest cumulative case rate
+
+sites <- data.frame(latitude = c(35.4044, 34.5258, 34.1065, 33.6754, 29.8688,
+                                  28.4117, 29.3343, 28.7410, 28.9089, 29.7525),
+                    longitude = c(-101.9805, -100.1703, -101.8892, -101.7980, -101.1617,
+                                  -99.8125, -99.8125, -100.3498, -97.8722, -94.6864))
+#Potter: 35.4044° N, 101.9805° W
+#Childress: 34.5258° N, 100.1703° W
+#Hale: 34.1065° N, 101.8892° W
+#Lubbock: 33.6754° N, 101.7980° W
+#Val Verde: 29.8688° N, 101.1617° W
+#Dimmit: 28.4117° N, 99.8125° W
+#Uvalde: 29.3343° N, 99.8125° W
+#Maverick: 28.7410° N, 100.3498° W
+#Karnes: 28.9089° N, 97.8722° W
+#Chambers: 29.7525° N, 94.6864° W
+
 map_rates <- ggplot(merge_shp_vac) +
-  geom_sf(aes(fill = CumCaseRatePer100)) +
+  geom_sf(aes(fill = CumCaseRatePer100), colour = "white", size = 0.05) +
+  geom_point(data = sites, aes(x = longitude, y = latitude), size = 2, shape = 23, fill = "black") +
+  geom_sf_text(data = subset(merge_shp_vac, CumCaseRatePer100 >= 19.1), aes(label = COUNTY), size = 4.5, nudge_y = -0.13) +
   ggtitle("Cumulative Confirmed Case Rate/100 Persons, by County, as of 10/3/2021") +
-  geom_sf_text(data = subset(merge_shp_vac, CumCaseRatePer100 >= 191), aes(label = COUNTY), size = 2.5) +
   scale_fill_gradient(low = "light blue", high = "red", name = "Cumulative Confirmed Case Rate/100 Persons", labels = comma)
@@
 map_rates
-#Very different results, hmm...let's summarize county-level characteristics
 
@@
 ggsave(filename = figure_file_map_rates, plot=map_rates, height = 12, width = 15) 
+```
+
+Epidemic Curve, by County
+```{r}
+#Too many counties for facet_wrap, narrow down to top 10 counties with highest case burden
+NewCasesTime_long %>%
+  group_by(County) %>%
+  summarize(Sum = sum(NewCases)) %>%
+  arrange(desc(Sum))
+
+NewCasesTime_long_subset <- NewCasesTime_long %>%
+  filter(County %in% c("Dimmit", "Hale", "Uvalde", "Maverick", "Karnes", "Childress", "Lubbock", "Val Verde", "Potter", "Chambers"))
+
+epicurve_county <- ggplot(NewCasesTime_long_subset, aes(x=Date, y = NewCases)) +
+  geom_bar(stat = "identity", fill="darkblue") +
+  scale_x_date(date_breaks = "months" , 
+               date_labels = "%b-%y", 
+               limits = c(min(NewCasesTime_long$Date), max = max(NewCasesTime_long$Date))) + 
+  scale_y_continuous(limits = c(min=0, max = 750)) +
+  facet_wrap(~County, nrow = 2) + #2 rows 
+  theme(axis.text.x = element_text(angle=50, hjust=1)) +
+  labs(title = "Epidemic Curve", 
+       subtitle = "Subset to counties with the highest case burden",
+       y = "Cumulative Case Rate", x = "")
+
+epicurve_county
+
+#Save figure
+figure_file_epicurve_county = here("results","epicurve_county.png")
+ggsave(filename = figure_file_epicurve_county, plot=epicurve_county, height = 5, width = 15) 
 ```
@@
 saveRDS(table_county, file = save_data_location6)
+
+library(vtable)
+summarytable <- sumtable(merge_shp_vac_df_c, summ = c('mean(x)', 'min(x)', 'max(x)'),
+         title="", out = "return")
+
+#Save table
+save_data_location17 <- here::here("results","summarytable.rds")
+saveRDS(summarytable, file = save_data_location17)
 ```
 
-#Association between county-level characteristics and # confirmed cases 
+Association between county-level characteristics and # confirmed cases 
 ```{r}
-p1 <- ggscatter(merge_shp_vac_subset, x="PercentHispanic", y="CumCaseRatePer100",
-          add = "reg.line", conf.int = TRUE,
-          cor.coef = TRUE, cor.method = "spearman")
-p2 <- ggscatter(merge_shp_vac_subset, x="Percent65Plus", y="CumCaseRatePer100",
-          add = "reg.line", conf.int = TRUE,
-          cor.coef = TRUE, cor.method = "spearman")
-p3 <- ggscatter(merge_shp_vac_subset, x="PercentMale", y="CumCaseRatePer100",
-          add = "reg.line", conf.int = TRUE,
-          cor.coef = TRUE, cor.method = "spearman")
-p4 <- ggscatter(merge_shp_vac_subset, x="PercentBlack", y="CumCaseRatePer100",
-          add = "reg.line", conf.int = TRUE,
-          cor.coef = TRUE, cor.method = "spearman")
-p5 <- ggscatter(merge_shp_vac_subset, x="PercentAsian", y="CumCaseRatePer100",
-          add = "reg.line", conf.int = TRUE,
-          cor.coef = TRUE, cor.method = "spearman")
-p6 <- ggscatter(merge_shp_vac_subset, x="PercentMedCondition", y="CumCaseRatePer100",
-          add = "reg.line", conf.int = TRUE,
-          cor.coef = TRUE, cor.method = "spearman")
-p7 <- ggscatter(merge_shp_vac_subset, x="AREA_SQKM", y="CumCaseRatePer100",
-          add = "reg.line", conf.int = TRUE,
-          cor.coef = TRUE, cor.method = "spearman")
+fit1 <- lm(`Confirmed Cases` ~ PercentHispanic, data = merge_shp_vac_subset, offset = Population)
+summary(fit1) #Adjusted R-Squared = 0.56
 
-p_combined <- plot_grid(p1, p2, p3, p4, p5, p6, p7)
+fit2 <- lm(`Confirmed Cases` ~ Percent65Plus, data = merge_shp_vac_subset, offset = Population)
+summary(fit2) #Adjusted R-Squared = 0.56
 
-# Save figure
-figure_file_association = here("results","p_combined.png")
-ggsave(filename = figure_file_association, plot=p_combined, height = 5, width = 15.5)
+fit3 <- lm(`Confirmed Cases` ~ PercentMale, data = merge_shp_vac_subset, offset = Population)
+summary(fit3) #Adjusted R-Squared = 0.56
 
+fit4 <- lm(`Confirmed Cases` ~ PercentMedCondition, data = merge_shp_vac_subset, offset = Population)
+summary(fit4) #Adjusted R-Squared = 0.56
+
+fit5 <- lm(`Confirmed Cases` ~ PercentAsian, data = merge_shp_vac_subset, offset = Population)
+summary(fit5) #Adjusted R-Squared = 0.57
+
+fit6 <- lm(`Confirmed Cases` ~ PercentBlack, data = merge_shp_vac_subset, offset = Population)
+summary(fit6) #Adjusted R-Squared = 0.56
+
+fit7 <- lm(`Confirmed Cases` ~ AREA_SQKM, data = merge_shp_vac_subset, offset = Population)
+summary(fit7) #Adjusted R-Squared = 0.56
+#Why are they all about the same R-squared? I shouldn't use correlation tests (spearman, pearson) here because I need the offset
+
+#Cumulative Case Rate
+fit_rate <-  lm(sqrt(CumCaseRatePer100) ~ AREA_SQKM, data = merge_shp_vac_subset)
+summary(fit_rate) #Why is the R-squared so low when I try this (with or without transformation)?
+
+cor.test(merge_shp_vac_subset$CumCaseRatePer100, merge_shp_vac_subset$AREA_SQKM, method = "pearson")
+
+#I am so confused...
+#Cumulative Case Rate
+ggplot(merge_shp_vac_subset, aes(y = log(CumCaseRatePer100), x = AREA_SQKM)) +
+       geom_point() +
+       geom_smooth(method = "lm")
+
+#Confirmed Cases
+ggplot(merge_shp_vac_subset, aes(y = log(`Confirmed Cases`), x = AREA_SQKM)) +
+       geom_point() +
+       geom_smooth()
+hist(merge_shp_vac_subset$AREA_SQKM) #is there just not enough range?
+#When I don't add lm, looks like a polynomial, but fit was worse when added ^2 or ^2 and ^3
+
+#Initially, I fit confirmed case against population numbers (# 65+) instead of percentages ((%65+) and the R-squared was like 0.99, so I think this is confounding because it's just getting at population size
+
+#Confirmed Case
+hist(log(merge_shp_vac_subset$`Confirmed Cases`)) #LOG TRANSFORMATION GIVES THE OUTCOME A NORMAL DISTRIBUTION! Since sample size is pretty small (254), law of large numbers doesn't apply.
+shapiro.test(log(merge_shp_vac_subset$`Confirmed Cases`))
+#Did try fitting log(Cofirmed Cases), but r-squared and AIC was worse
+
+#Cumulative Case Rate
+hist(merge_shp_vac_subset$CumCaseRatePer100)  
+shapiro.test(sqrt(merge_shp_vac_subset$CumCaseRatePer100)) #SQUARE ROOT TRANSFORMATION GIVES THE OUTCOME A NORMAL DISTRIBUTION
 ```
@@
 ```{r}
-# p1 <- ggplot(merge_shp_vac_subset, aes(x=Population, y=`Confirmed Cases`)) +
-#   geom_point() +
-#   geom_smooth(method='lm') + #add linear regression line 
-#   #stat_regline_equation(label.y = 570000, aes(label = ..eq.label..)) +
-#   stat_regline_equation(aes(label = ..rr.label..)) +
-#   scale_x_continuous(labels = comma) + #remove scientific notation
-#   scale_y_continuous(labels = comma) 
+# p1 <- ggscatter(merge_shp_vac_subset, x="PercentHispanic", y="CumCaseRatePer100",
+#           add = "reg.line", conf.int = TRUE,
+#           cor.coef = TRUE, cor.method = "spearman")
 #
-# p2 <- ggplot(merge_shp_vac_subset, aes(x=PercentMale, y=CumCaseRatePer100)) +
-#   geom_point() +
-#   geom_smooth(method='lm') + #add linear regression line 
-#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
-#   stat_regline_equation(aes(label = ..rr.label..)) +
-#   scale_x_continuous(labels = comma) + #remove scientific notation
-#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
+# p2 <- ggscatter(merge_shp_vac_subset, x="Percent65Plus", y="CumCaseRatePer100",
+#           add = "reg.line", conf.int = TRUE,
+#           cor.coef = TRUE, cor.method = "spearman")
+# p3 <- ggscatter(merge_shp_vac_subset, x="PercentMale", y="CumCaseRatePer100",
+#           add = "reg.line", conf.int = TRUE,
+#           cor.coef = TRUE, cor.method = "spearman")
+# p4 <- ggscatter(merge_shp_vac_subset, x="PercentBlack", y="CumCaseRatePer100",
+#           add = "reg.line", conf.int = TRUE,
+#           cor.coef = TRUE, cor.method = "spearman")
+# p5 <- ggscatter(merge_shp_vac_subset, x="PercentAsian", y="CumCaseRatePer100",
+#           add = "reg.line", conf.int = TRUE,
+#           cor.coef = TRUE, cor.method = "spearman")
+# p6 <- ggscatter(merge_shp_vac_subset, x="PercentMedCondition", y="CumCaseRatePer100",
+#           add = "reg.line", conf.int = TRUE,
+#           cor.coef = TRUE, cor.method = "spearman")
+# p7 <- ggscatter(merge_shp_vac_subset, x="AREA_SQKM", y="CumCaseRatePer100",
+#           add = "reg.line", conf.int = TRUE,
+#           cor.coef = TRUE, cor.method = "spearman")
 # 
-# p3 <- ggplot(merge_shp_vac_subset, aes(x=PercentBlack, y=CumCaseRatePer100)) +
-#   geom_point() +
-#   geom_smooth(method='lm') + #add linear regression line 
-#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
-#   stat_regline_equation(aes(label = ..rr.label..)) +
-#   scale_x_continuous(labels = comma) + #remove scientific notation
-#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
-# 
-# p4 <- ggplot(merge_shp_vac_subset, aes(x=PercentAsian, y=CumCaseRatePer100)) +
-#   geom_point() +
-#   geom_smooth(method='lm') + #add linear regression line 
-#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
-#   stat_regline_equation(aes(label = ..rr.label..)) +
-#   scale_x_continuous(labels = comma) + #remove scientific notation
-#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
-# 
-# p5 <- ggplot(merge_shp_vac, aes(x=PercentHispanic, y=CumCaseRatePer100)) +
-#   geom_point() +
-#   geom_smooth(method='lm') + #add linear regression line 
-#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
-#   stat_regline_equation(aes(label = ..rr.label..)) +
-#   scale_x_continuous(labels = comma) + #remove scientific notation
-#   scale_y_continuous(labels = comma, limits = c(0,30))
-# 
-# p6 <- ggplot(merge_shp_vac_subset, aes(x=Population_Density, y=CumCaseRatePer100)) +
-#   geom_point() +
-#   geom_smooth(method='lm') + #add linear regression line 
-#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
-#   stat_regline_equation(aes(label = ..rr.label..)) +
-#   scale_x_continuous(labels = comma) + #remove scientific notation
-#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
-# 
-# p7 <- ggplot(merge_shp_vac_subset, aes(x=Percent65Plus, y=CumCaseRatePer100)) +
-#   geom_point() +
-#   geom_smooth(method='lm') + #add linear regression line 
-#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
-#   stat_regline_equation(aes(label = ..rr.label..)) +
-#   scale_x_continuous(labels = comma) + #remove scientific notation
-#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
-# 
-# p8 <- ggplot(merge_shp_vac_subset, aes(x=PercentVaccinated_1dose, y=CumCaseRatePer100)) +
-#   geom_point() +
-#   geom_smooth(method='lm') + #add linear regression line 
-#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
-#   stat_regline_equation(aes(label = ..rr.label..)) +
-#   scale_x_continuous(labels = comma) + #remove scientific notation
-#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
-# 
-# p9 <- ggplot(merge_shp_vac_subset, aes(x=PercentVaccinated_full, y=CumCaseRatePer100)) +
-#   geom_point() +
-#   geom_smooth(method='lm') + #add linear regression line 
-#   #stat_regline_equation(label.y = 550000, aes(label = ..eq.label..)) +
-#   stat_regline_equation(aes(label = ..rr.label..)) +
-#   scale_x_continuous(labels = comma) + #remove scientific notation
-#   scale_y_continuous(labels = comma, limits = c(0, 30)) 
-# 
-# p_combined <- plot_grid(p1, p2, p3, p4, p5, p6, p7, p8, p9) 
-# 
-# #Save figure
-# figure_file_association = here("results","p_combined.png")
-# ggsave(filename = figure_file_association, plot=p_combined, height = 5, width = 15.5)
-
+# p_combined <- plot_grid(p1, p2, p3, p4, p5, p6, p7) 
 ```

### 2021-11-29 products/manuscript/Manuscript.Rmd
Project part 5

@@
 ---
-title: 'MADA Data Analysis Project: COVID'
-author: "Andreas Handel"
+title: 'MADA Data Analysis Project: COVID-19 Case Burden, Texas'
+author: "Gabriella Veytsel"
 date: '`r Sys.Date()`'
@@
   bookdown::word_document2:
-    number_sections: yes
+    number_sections: no
     toc: no
@@
 # Summary/Abstract
-Understanding the patterns in COVID-19 burden at the county-level can inform targeted public health interventions and resource allocation.
+There is substantial heterogeneity in population demographics and health status within states; therefore, a model is needed with a more granular resolution.
+In this study, we focus on the state of Texas, aiming to characterize the inter-county variability and identify important predictors of COVID-19 case burden.
+We located counties with the highest case burdens, visualized their trends over time, and identified the proportion of Hispanic individuals as the most important feature in our final machine learning model.
+Given limited resources, decisions for COVID-19 testing, monitoring, and interventions must be data-driven: understanding the patterns in COVID-19 burden at the county-level can inform targeted public health interventions and resource allocation.
 
 # Introduction 
-Given limited resources, decisions for COVID-19 testing, monitoring, and interventions must be data-driven.
-We aim to characterize the inter-county variability in COVID-19 burden and identify important predictors to help guide public health response and resource allocation, such as vaccine distribution, deployment of mobile testing units, health messaging/education, wastewater monitoring, and hospital equipment.
-Utilizing county-level risk factors in decision-making, in addition to case counts, hospitalizations, and deaths, may allow for a more robust public health response.
-
-## General Background Information
-Underreporting in cases, due to county-level variation in testing availability and individual-level differences in healthcare access and health-seeking behavior, is a standard consideration in disease reporting.
-With the significant proportion of asymptomatic cases and issues in testing capacity during the early phase, it is a particularly important challenge in the current COVID-19 epidemic.
+As of November 19, 2021, the unprecedented coronavirus disease 2019 (COVID-19) pandemic, caused by severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2), has accounted for at least 47,030,792 cases and 759,388 deaths in the United States alone [@WHO-2021].
+We hypothesized that the proportion of some high-risk groups may be importants predictors for COVID-19 case burden.
+According to the CDC, age is the strongest risk factor for severe COVID-19 morbidity; additionally, the risk of severe COVID-19 increase as the number of underlying medical conditions increase in an individual [@CDC-risk].
+Communities of color in the United States experience structural, social, and environmental inequities that increase risk for infection and severe illness [@Bhala-2020] and have experienced substantial barriers to and disparities in COVID-19 testing [@NPR-2020].
+Given the health disparities for high-risk and vulnerable groups at the individual level, we sought to examine the associations between county-level demographic data and cumulative COVID-19 case rate.
+These findings may help guide public health response and resource allocation, such as vaccine distribution, deployment of mobile testing units, health messaging/education, wastewater monitoring, and medical resources (i.e., personal protective equiment and ICU beds).
 
 ## Description of data and data source
-* Data from the Texas Health Department: https://dshs.texas.gov/coronavirus/AdditionalData.aspx
-    + Cases.and.Fatalities: case count data at the county-level, total 
-    + Trends: new cases at the state-level, by date 
-    + NewCasesTime: new cases at the county-level, by date
-    + Demographics: sex, race/ethnicity
-    + Vaccination: 1-dose and fully vaccinated population, population 65+ from 2019 Census population estimates, population 16-64 any medical condition*
-
-* Shapefile of Texas with county boundaries: https://www.depts.ttu.edu/geospatial/center/TexasGISData.html
-    + Includes data on county population and area to calculate county population density
+Cumulative confirmed COVID-19 case counts at the county-level, confirmed COVID-19 case counts at the county- and state-level over time, and 2019 county-level population estimates were obtained from the Texas Department of State Health Services [@Texas-dashboard] on October 3rd, 2020.
+The shapefile of Texas with county borders and data on county size was obtained from the Texas Tech University Center for Geospatial Technology [@Texas-GIS].
 
 ## Questions/Hypotheses to be addressed
-* Investigate the spatial heterogeneity of COVID-19 incidence within Texas over time at the county-level.
-* Evalauate the associations between county-level characteristics (e.g., population density, population demographics, and vaccination coverage) and COVID-19 burden.
+1. Investigate the spatial heterogeneity of COVID-19 cumulative incidence within Texas.
+2. Investigate county-level incidence of COVID-19 within Texas over time.
+3. Evalauate the associations between county-level characteristics and COVID-19 burden within Texas.
 
 # Methods and Results
-Cumulative numbers of COVID-19 cases and 2019 population estimates for each county were obtained from the Texas Department of Health.
-As counties containing major cities (e.g., Houston, Dallas, Fort Worth) had the highest total cases, cumulative case rates per 100 persons were calculated and mapped to visualize spatial heterogeneity across Texas counties.
-Epidemic curves were constructed to visualize new cases over time at the state- and county- levels (subset to 10 counties with the highest total case burden).
-County-level risk factors included in our analysis were: age (65 and older), gender (male), race/ethnicity (Black, Asian, and Hispanic), and any medical condition that increases risk of COVID*.
+The cumulative case rate across Texas was 11.7 cases per 100 persons, calculated by dividing the number of confirmed cases as of October 3, 2020 (n = 3,394,815) by the 2019 population estimate of 29,001,602 people.
+The cumulative case rate per county ranged from 2.6 (King County, TX) to 31.3 cases (Dimmit County, TX) per 100 persons.
+On average, the counties were 1% Asian, 6.2% Black, 50.9% male, 18.5% over 65 years of age, and 29.7% had at least one medical condition.
+Additionally, on average, the counties had a 2019 population of 114,180 persons and an area of 2,739 square kilometers.
+As expected, counties with major cities had the highest total cases; in order to compare across counties with different population sizes, cumulative case rates per 100 persons were mapped to visualize spatial heterogeneity across Texas.
+Additionally, epidemic curves were constructed to visualize new cases over time at the state- and county- levels (subset to 10 counties with the highest cumulative case rates).
+In Texas, the three waves of the epidemic appear to peak around the end of July, beginning of January, and end of September.
+These trends are mostly reflected in the county-level maps, perhaps with a few deviations due to differences in testing rates and scale of interventions.
+County-level risk factors included in our analysis were: age (% 65 and older), gender (% male), race/ethnicity (% Black, Asian, and Hispanic), and any medical condition that increases risk of severe illness from COVID-19*.
+Population density and population size were found to be significantly correlated (Pearson's correlation coefficient: 0.94, 95% CI: 0.93, 0.95); since our models include a population component in its rate calculation, we chose to drop population density in favor of county area.
 
-To estimate the association between county-level risk factors and cumulative case incidence, the following models were developed: Poisson regression, negative binomial regression, multivariable linear regression, and a generalized mixed effects model.
-For the Poisson model, the assumption that the mean and variance of number of cumulative COVID-19 cases were approximately equal was violated; additionally, there was evidence of severe overdispersion.
+To estimate the association between county-level risk factors and cumulative case rate, the following models were developed: Poisson regression, negative binomial regression, generalized mixed effects model, and multivariable linear regression.
+For the Poisson model, the assumption that the mean and variance of the outcome, cumulative COVID-19 case counts, were approximately equal was violated; additionally, there was evidence of severe overdispersion.
 Ultimately, this model was rejected as an appropriate model choice.
 The negative binomial regression model was included to relax the assumption of mean-variance equivalence and, accordingly, had superior performance.
-Several diagnostic plots (fitted vs.
+Additionally, the generalized mixed effects model incorporated county as a random effect in order to capture the inter-county variation in testing rates.
+These three aformontioned models fit case count as the outcome of interest offset by the natural logarithm of the 2019 county-level population estimates in order to estimate cumulative incidence rate ratios.
+On the other hand, we assigned cumulative case rates as our continuous outcome of interest in a linear regression model, utilizing the Tidymodels framework in R.
+We utilized two machine learning methods for variable selection and hyperparameter tuning, regularization, and optimization: 1) a penalized linear regression model (LASSO), which enacts a regularization penalty on coefficients that shrinks the impact of or drops irrelevant predictors from the model, and 2) a random forest model, which uses a bagging method to build an ensemble of decision trees (MADA 2021).
+Before fitting these models, we fit a simple linear regression model to each potential predictor, assessing the degree to which the response variable is explained by the predictor via r-squared.
+All analyses were conducted in R version 4.1.0 (2021-05-18) [@R-base].
+
+To ensure that our predictors were informative in all our models, we built a null model as a baseline for model performance (RMSE on training data = 4.30).
+For model evaluation, we inspected model performance metrics (root mean square error (RMSE) and r-squared) and diagnostic plots.
+Fitted vs.
 residual plots, normal QQ plot, scale-location, and residuals vs.
-leverage) were inspected to test if linear regression model assumptions were met.
-Indeed, the assumptions of linearity of the predictor and outcome, constant variance of residuals, normality of residuals, and independence of residuals error terms of the multivariable linear regression were accepted.
+leverage were compared between models for the assumption of linearity of the predictor and outcome, constant variance of residuals, normality of residuals, and independence of residuals error terms.
+For our linear regression model that was built in TidyModels, we were also able to split the dataset into training and testing data sets to perform cross-validation.
+LASSO removed the % male feature and applied a penality of 0.1 to remaining predictors.
+While the random forest model does not remove predictors, the VIP package in R [@VIP-R] also selected % male as the second least important variable.
+The best model for each approach was chosen based on RMSE; finally, the model performance and diagnostic plots for the best LASSO model and random forest model were compared.
+Since the model performance for the best LASSO model (RMSE = 3.82) and the best random forest model (RMSE = 3.77) were very similar for the training data, we selected the random forest model as our final model and proceeded to conducting 5-fold cross-validation on the test dataset one single time.
+Compared to the model performance of our null model on testing data (RMSE = 3.81), we can see that our final model had a slightly superior performance metric (RMSE = 3.32).
 
-We used the TidyModels framework in R to build a penalized linear regression model (LASSO) and Random Forest regression model.
-LASSO removed the proportion male variable and applied a penality of 0.1 to remaining predictors.
-While the Random Forest model does not remove predictors, the VIP package in R also selected PercentMale as the least important variable.
-The best model for each approach was chosen based on RMSE and diagnostic plots, which was very similar (LASSO RMSE = 3.82, RF RMSE = 3.76).
-We selected the LASSO model as the final model and conducted 5-fold cross-validation on the test dataset one single time.
+Our final model identified the following predictors as the most important, in order: % Hispanic, % 65+, % Medical Condition, % Asian, % Black, % Male, and area.
+Still, the low r-squared (rsq = 0.244) and the diagnostic plots indicate that model results should not be over-interpreted.
 
-*People 16 years of age and older with at least one chronic medical condition that puts them at increased risk for severe illness from the virus that causes COVID-19.
+*According to the Texas Department of Health: "People 16 years of age and older with at least one chronic medical condition that puts them at increased risk for severe illness from the virus that causes COVID-19.
 The data are model-based estimates from CDC using self-reported data from the Behavioral Risk Factor Surveillance System and U.S.
 census population data.
-The methodology for generating these estimates are detailed in this MMWR: https://www.cdc.gov/mmwr/volumes/69/wr/mm6929a1.htm.
-
-## Data import and cleaning
+The methodology for generating these estimates are detailed in this MMWR: https://www.cdc.gov/mmwr/volumes/69/wr/mm6929a1.htm".
 
@@
 
-County-level characteristics across Texas
+Map of Texas, highlighting counties with the highest COVID-19 case burden
 
-```{r summarytable,  results = 'asis', echo=FALSE, message = FALSE, warning = FALSE}
+```{r map_rates, fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/map_rates.png")
+
+```
+
+Average characteristics across Texas
+
+```{r summarytable, echo=FALSE, message = FALSE, warning = FALSE}
 library(tidyverse)
@@
 
-Map of Texas, highlighting counties with the highest COVID-19 case burden
+County-level characteristics across Texas, subset to counties with the highest COVID-19 case burden
 
-```{r map_rates,  fig.cap='', echo=FALSE}
-knitr::include_graphics("../../results/map_rates.png")
-```
-
-County-level characteristics for Texas counties with the highest COVID-19 case burden
-
-```{r table_county,  results = 'asis', echo=FALSE, echo=FALSE, message = FALSE, warning = FALSE}
+```{r table_county, echo=FALSE, message = FALSE, warning = FALSE}
 library(tidyverse)
 
-resulttable=readRDS("../../results/table_county.rds")
-knitr::kable(resulttable, caption = '') %>%
+resulttable_county=readRDS("../../results/table_county.rds")
+knitr::kable(resulttable_county, caption = '') %>%
   kableExtra::kable_classic(full_width = F, html_font = "Times New Roman")
@@
 
-Association between county-level characteristics and total confirmed COVID-19 cases.
-Spearman's correlation coefficient and regression line
+## Full Analysis
 
-```{r associations,  fig.cap='', echo=FALSE}
-knitr::include_graphics("../../results/p_combined.png")
+Variable Importance, Random Forest Model
+
+```{r rf_vip,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/rf_vip.jpeg")
 ```
 
-## Full analysis
+# Discussion
+
+## Summary and Interpretation
+Our final model selected the percent of Hispanic individuals as the most important predictor; in fact, it had at least 2.5 times more of an impact compared to the second most important model feature, % 65+.
+Of note, 4 out of the 10 counties with the highest case burden in the state border Mexico and have a population that is 47-95% Hispanic (compared to the state average of 37%).
+Resource allocation must be targeted; for example, these counties have a high proportion of first-dose vaccination (54-89%).
+Ramping up health communication (with Spanish translation) around prevention, wastewater monitoring, and allocation of medical resources may be more appropriate for these specific communities.
+
+Underreporting in cases, due to county-level variation in testing availability and individual-level differences in healthcare access and health-seeking behavior, is a standard consideration in disease reporting.
+With the significant proportion of asymptomatic cases and issues in testing capacity during the early phase, it is a particularly important challenge in the current COVID-19 epidemic.
+We believe that utilizing county-level risk factors, in addition to counts of cases, hospitalizations, and deaths, may allow for a more robust public health response.
+
+## Strengths and Limitations
+There are many potentially important predictors that were not considered in this study, such as county-level proportions for urban/rural, household size, and socio-economic status.
+Outcomes should be adjusted for testing rates to more fairly compare across counties.
+While the mixed effects model had the best performance metric of all models (RMSE = 2.82), we're unable to assess overfitting without performing cross-validation; thus, we did not move forward on this model.
+Additionally, we did not add a spatial component to any model due to the complexity of incorporating case counts for out of state (and country) counties that border Texas and may be important drivers of transmission; this potentially may have improved model fit.
+Comorbidities here are aggregated; however, some have more evidence than others.
+
+## Conclusions
+This study identified counties with the highest COVID-19 burden and important predictors for county-level COVID-19 cumulative case rates.
+
+## Supplementary Figures
 Negative Binomial Model Evaluation
@@
 
+```{r LASSO_tuning,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/LASSO_tuning.jpeg")
+```
+
+```{r LASSO_regularization,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/LASSO_regularization.jpeg")
+```
+
 Random Forest Model Evaluation
@@
 
-# Discussion
+Final Model (Random Forest) on Test Data
 
-## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+```{r rf_final_pred_vs_obs,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/rf_testing_pred_vs_obs.jpeg")
+```
 
-## Strengths and Limitations
-There are many potentially important predictors that were not considered in this study, such as county-level proportions for urban/rural, household size, and socio-economic status.
-
-## Conclusions
-_What are the main take-home messages?_
-
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-This paper [@Leek2015a] discusses types of analyses.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
+```{r rf_final_residuals,  fig.cap='', echo=FALSE}
+knitr::include_graphics("../../results/rf_testing_residuals.jpeg")
+```
 

### 2021-11-29 code/analysis_code/analysisscript.R

@@ -11,6 +11,7 @@
 
 #Load packages
 library(tidyverse)
+library(dplyr)
 library(tidymodels)
 library(poissonreg)
 library(lme4)
@@ -21,6 +22,7 @@ library(performance) #overdispersion
 library(AER) #overdispersion
 library(MASS) #negative binomial
 library(vip)
+library(skimr)
 
 #Load data
 save_data_location1 <- here::here("data","processed_data","merge_shp_vac_df_c.rds")
@@ -37,25 +39,41 @@ cleandata_Vaccination <- readRDS(save_data_location4)
 NewCasesTime_long <- readRDS(save_data_location5)
 cleandata_Demographics <- readRDS(save_data_location6)
 
-combined <- combined %>%
+head(combined)
+sum(combined$Population) #29,001,602 people
+sum(combined$`Confirmed Cases`) #3,394,815
+sum(combined$`Confirmed Cases`)/sum(combined$Population) * 100 #11.7 cases per 100 persons
+summary(combined$CumCaseRatePer100) #cumulative case rate per county ranged from 2.6 to 31.3 cases per 100 persons
+
+skim(combined)
+
+combined1 <- combined %>%
   dplyr::select(-COUNTY)
 
-#Is cumulative case rate normally distributed? No
-hist(combined$CumCaseRatePer100, main = "Histogram of Cumulative Case Rate per 1,000", xlab = "Cumulative Case Rate per 1,000")
-ggdensity(combined$CumCaseRatePer100, main = "Density Plot of Cumulative Case Rate per 1,000", xlab = "Cumulative Case Rate per 1,000")
-ggqqplot(combined$CumCaseRatePer100) #QQ plot
+#Population size and Population density correlated? Yes
+cor(combined$Population, combined$Population_Density, method = c("pearson", "kendall", "spearman"))
+cor.test(combined$Population, combined$Population_Density, method=c("pearson", "kendall", "spearman"))
 
-shapiro.test(combined$CumCaseRatePer100) #p-value = 5.8E-7 
+#Population and area? No
+cor(combined$Population, combined$AREA_SQKM, method = c("pearson", "kendall", "spearman"))
+cor.test(combined$Population, combined$AREA_SQKM, method=c("pearson", "kendall", "spearman"))
+
+#Is cumulative case rate normally distributed? No. 
[…]

### 2021-11-29 code/processing_code/processingscript.R

@@ -141,6 +141,9 @@ merge_shp_vac <- merge_shp %>%
   mutate(PercentBlack = NH_Black_Total /  Total * 100) %>%
   mutate(PercentMale = Total_Male / Total * 100)
 
+sum(merge_shp_vac$Population65Plus)/sum(merge_shp_vac$Total) #.129
+mean(merge_shp_vac$Percent65Plus) #18.5
+
 # save data as RDS
 # location to save file
 save_data_location1 <- here::here("data","processed_data","merge_shp_vac.rds")

### 2021-12-03 project_review-jmartin.Rmd
Project Review - JM

---
title: Project Review
date: "`r file.mtime(knitr::current_input())`"
#bibliography: ../media/references.bib
output: 
  html_document:
    toc_depth: 3
    number_sections: true
---

# Overview

Title of project:"MADA Data Analysis Project: COVID-19 Case Burden, Texas"

Name of project author(s): Gabriella Veytsel

Name of project reviewer: Joe Martin

# Specific project content evaluation

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

Goal of this project is to predict where COVID-19 outbreaks are likely to occur given data from counties in Texas. It is clear why the project was undertaken, but I would like to read more about the background. Have other researchers tried similar approaches? What worked and what didn't work?

### Summary assessment

* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

Questions and hypotheses are very clear. Goal is to investigate hereogeneity of COVID-19 in Texas, investigate cases ate the county level, and evaluate associations between county demographics and COVID cases. It would be helpful to have a little more information about some of the variables being tested in this section.

### Summary assessment

* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

Sources for the data are provided and the type of data available is intuitive - especially when reviewing the first paragraph in the Methods section. Like previous comment, it would be helpful to have a brief description of variables or types of variables available (ex. how does the health dataset present medical conditions? Does it list conditions for individuals? Does it just provide a count of medical conditions?). It would also be helpful to have more information in the readme file. The link to the Texas DHHS dashboard is helpful, but also confusing when confronted with all of the different data links.

### Summary assessment

* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

Data cleaning/wrangling script is excellent. All steps were clearly defined and documented. I was really impressed by the figures in your exploratory analysis. I think all of these graphics did an excellent job telling a story. I especially liked the state maps. Only feedback I have for your exploratory is that I do wish there was a little more commentary so I could understand how your figures and tables relate to your paper and how you plan to use them.

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

Your analysis is easy to follow and your comments clearly discuss your and your actions. It seems like you tested several different models and did a good job of reviewing correlations and distributions in your data.

### Summary assessment

* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

The figures in your paper are high quality and do a good job of illustrating your research. It might help to have a little more explanation accompanying your figures to interpret what it should mean to the reader. For the supplementary figures, I'm not sure if you should take them out of the manuscript file, but I would think that section would improve if you could find a way to display them side-by-side instead of just having them vertical.

### Summary assessment

* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

You did a good job of discussing the strengths and limitations of your models. I think there is more you could write in your "Strengths and Limitations" section. For example, you mentioned using cross-validation in your methods, but then your wrote that you are unable to assess overfitting without performing cross-validation. Did your models that used cross-validation not run? Was there another reason you this wasn't possible? What models/resampling methods would you recommend trying next?

### Summary assessment

* minor parts wrong, missing or unclear

## Further comments

_Hi Gabby,_

_You did such an excellent job creating figures for your project! I'm especially impressed with your maps. It seems like you have many more useful figures related to your models, as well. The only thing I could think about changing would be the size and display of your tables. I'm not sure how to do this, but one suggestion I would have is to highlight individual values that are significant in your County table. It would also help to update the headers on this table._

_Good luck on the rest of your project!_

_-Joe_

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

I found a couple of empty or outdated readme files, as well as unused files and folders (for example /products/slides/...). Other than that, your repository seems to look good. It might be useful to list the files needed to run in your main readme file.

### Summary assessment

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

There were a couple of files I found that were missing comments where I thought they would have been useful, like your exploration script. There were certain places in your manuscript where I wanted more information about your analysis, as well. Other than that, your files have good, clear documentation.

### Summary assessment

* decently documented with some gaps

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

There isn't documentation with instructions to reproduce your project, but I didn't run into any issues running your code. Like I mentioned earlier, it might be useful to include a list of the necessary files to run in your main readme page.

### Summary assessment

* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

It seems like you were very thorough in your project. Again, I'm impressed by all of your figures and tables and it looks like you were able to run several different models which performed progressively better. I think you address your questions/hypotheses in "Methods and Results," but you have room to expand on them in your Discussion

### Summary assessment

* strong level of thoroughness

## Further comments

_Once again, good job on your project! I'm looking forward to reading your final paper._

### 2021-12-04 Hammerton_Review.Rmd

---
title: "Project Review for Gabriella Veytsel"
author: "Savannah Hammerton"
date: "Fall 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Title of project: MADA Data Analysis Project: COVID-19 Case Burden, Texas

Name of project author(s): Gabriella Veytsel

Name of project reviewer: Savannah Hammerton

## Background, Context and Motivation

### Feedback and Comments

You did a good job giving general background of why you performed this analysis.

In your summary/abstract, I would recommend being more clear when you say "a model is needed with a more granular resolution." Since you haven't mentioned any outcomes/public health issues yet, I was a little confused as to what model you were referring to.

Additionally, in the introduction section, when you say

"According to the CDC, age is the strongest risk factor for severe COVID-19 morbidity; additionally, the risk of severe COVID-19 increase as the number of underlying medical conditions increase in an individual (Assessing risk factors for severe COVID-19 illness, 2020). Communities of color in the United States experience structural, social, and environmental inequities that increase risk for infection and severe illness (Bhala, 2020) and have experienced substantial barriers to and disparities in COVID-19 testing (In large texas cities, access to coronavirus testing may depend on where you live, 2020)",

it might help to have something describing how those two statements are related. Since the first statement refers to more physical attributes and the second refers to more social/environmental attributes, it almost made the reasoning for the analysis weaker to me. It made it seem like maybe the focus should have been on the age/underlying medical conditions, since those are what the CDC mentioned as important factors.

I liked that you ended your introduction with what public health interventions could be impacted by the outcome of your analysis; that's definitely a great thing to have in there for motivation.

### Summary assessment

* some contextualization and motivation

## Question description

### Feedback and Comments

I like that you have your questions explicitly listed. It might help to have some sort of statement prior to the questions themselves to re-introduce the overall, broader question you are trying to answer by answering these specific questions.

### Summary assessment

* question/hypotheses fully clear

## Data description

### Feedback and Comments

From the description of the data, I don't see anything about data that describes populations. "Population estimates" makes me think of population counts in each county. Adding something about demographic information or what exact predictors you're using could help.

### Summary assessment

* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis

### Feedback and Comments

I haven't seen anywhere so far that explicitly states what your predictors are, how you are looking for them, or checking their distributions. I found one table giving summaries of demographics, but everything else seems to be focused on your outcome. I'm a little confused about whether your predictor is just counties, or if you're specifically using demographic information as predictors as well.

Otherwise, you have a lot of exploration of the outcome, and some really nice visualizations! I like that you made sure to assess both disease counts and rates. Your maps and epidemic curves look nice.

### Summary assessment

* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

### Feedback and Comments

I liked that I could see how the ML models improved on the performance on the performance of the log binomial and poisson models. I saw that conducted some log transformation in those models; I would recommend trying that in your ML models as well.`step_log()` and 'step_normalize()` could both help your numeric predictors, but checking the distributions of all your predictors would tell you specifically which ones really need it.

I would add some of the process you went through in the analyses to the manuscript, as well as a little more detail on why you selected the random forest as your final model (i.e., was there any reason besides the RMSE that you chose that over the LASSO model?).

### Summary assessment

* defensible but not optimal analysis

## Presentation

### Feedback and Comments

Overall, I think the figures you present look really nice! For cohesiveness' sake, they could benefit from all having the same background/theme. As it is, some have a white/blank background while others have the grey/lined default ggplot background. If you use `theme_set(*insert the theme you like here*)` at the beginning of scripts where you create plots, it could help the consistency of appearance. The plot sizes are also pretty variable, and some could stand to be a little bigger (I had to squint, but to be fair, my eyesight is terrible). Adding a little more description to figure captions could help as well with comprehension of the what you're trying to show!

### Summary assessment

* results are presented ok, with room for improvement

## Discussion/Conclusions

### Feedback and Comments

Your strengths and limitations section has a lot if great points, and seems very comprehensive! I think you could use some more discussion. You've got a some good explanations of the final model, but discussing the other models you fit and how they performed, and maybe why they may have performed the way they did, could be a good addition here. Your conclusion section could also use some more text, specifically referring to the final model, since such a huge part of this project/class is selecting and fitting a final model and assessing performance.

### Summary assessment

* major parts of discussion missing or wrong

## Further comments

Some or all of the text directly in the Methods section could be distributed between the exploratory/full analysis sections. This would help the flow of the manuscript and allow the figures currently in those sections to be more easily understood in the context of what you were doing when you generated them.

# Overall project content evaluation

## Structure

### Feedback and Comments

Everything is easy to find and makes sense! Your products folder still has the slides, which don't seem to contain any information from you, so you could probably delete those. Your README files also help it determining what to run if there's any confusion.

### Summary assessment

* well structured

## Documentation

### Feedback and Comments

Your analysis script has good comments, especially as you respond to output of your models. Your processing and exploration scripts could use some more commentary; this could help give better context for your decisions in the analysis script!

### Summary assessment

* decently documented with some gaps

## Reproducibility

### Feedback and Comments

Everything ran with no problems! It was also clear what should be run and in what order!

### Summary assessment

* fully reproducible without issues

## Thoroughness

### Feedback and Comments

I think the project could use some additions, but I've already mentioned them (namely, more exploration of predictors, maybe some transformation of predictors in the ML models). The manuscript could also use some extra commentary, and you could maybe discuss whether/how you were able to address and answer your research questions.

### Summary assessment

* decent level of thoroughness

## Further comments

This is a really interesting project! I think with just a few tweaks/additions, it could be really strong! Great job.

### 2021-12-06 Project_Review_Mallis_Veytsel.Rmd
Add files via upload

# Overview

Title of project: MADA Data Analysis Project: COVID-19 Case Burden, Texas

Name of project author(s): Gabriella Veytsel

Name of project reviewer: Nicholas Mallis

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

Hi! I also used county data for my project so this looks somewhat familiar. My outcome was vaccination rate though. I think you do a great job laying out the the background and motivation for the study. I agree that aggregate data like this could help inform intervention/vaccination campaigns.

### Summary assessment 
* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

I think you did a great job laying this out. I'm impressed that you asked three questions that all take on quite a bit of work to answer! My one suggestion would be to move some of these questions up to the end of the introduction. I feel like that's usually how we would see it in a paper... It starts with background and ends with research question or question(s) in your case. It also might be good to have it in sentence form instead of bulleted out. I like how clear it looks right now, but if Dr. Handel is wanting us to have this as something mostly ready to submit, he might like to see it written out. But I could be wrong :|

### Summary assessment
* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

I think you did a good job laying out the sources and appropriately citing them.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments
Great job on the map and the EPI curve! I know both of those can be tricky and a lot of work. I also like how you highlight certain counties with the highest rates. I also like your use of the table1 package. I've used that before but for some reason I decided not to use it for mine and it was ugly. I didn't see it here, but did you plot histograms for continuous variables? That could be good to do in order to explore how each are distributed. With my project I did that for each variable and found some weird stuff ( a lot of 0's which were actually missing). It helped me figure out what was going on in the data a bit more. Could help you here, but it looks like you know good and well what's going on in your data already. Okay  well nevermind... now that I got to bottom of your document I see some histograms and normality tests : )

I don't see any mention of missingness or code that looks for missing data in each variable in your exploration script. That could be good to explore. Here is some code I used to explore that in my project...

colMeans(is.na(complete))

Here 'complete' is the name of the dataset. It returns each variable's proportion of missing.

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments
I'm impressed at 1) how well you explained your analysis and 2) the fact that I actually understood most of it : )  I think it's great that you tried so many different methods/models and went in depth on diagnostic plots and performance measures. In most epi papers, I feel like they have methods and results as two seperate sections, but I feel like with ML models that's almost impossible to do. I like how you collapsed both into one section. It walks the reader through your logic, analysis steps, and results nicely. I did my project with it split up into methods and results, but I might rework mine like yours here.

### Summary assessment
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments
As I said before, I'm really impressed with the figures you put together. My only suggestion on those would be to add footnotes to them citing the sources (which I know can be sort of annoying). I meant to do that to my project before Monday, but forgot. I (think) Dr. Handel would want to see stuff like that given that it's supposed to be ready for publication.

The tables look great! But one thing I notice is that a lot of the variable names are still in their coded form. He might want to see them written out without the '_'s and what not, but I could be wrong. Another idea would be to include full names in the footnote. And same thing goes for the tables, maybe add footnotes citing the source (something I also didn't do but planned to).

One last thing. On my end here, the title for table 2 came out like this..  (#tab:table_county). I think you could fix that pretty easy. It might have been a typo. Another note on table 2... Maybe try using the round() function when summarizing the data to get the results down to one or two digits after the decimal. It might make your table smaller and easier to read. When the numbers are higher like this, I don't think anything after one or two digits really adds much.

### Summary assessment
* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments
Based upon your logic and modeling strategies (and my new found understanding of ML methods), it seems to me that you have interpreted and explained the results of your study correct. One other limitation that you might want to mention could be the fact that some of your data comes from previous years and these metrics could have changed between 2019 and now.

### Summary assessment
* strong, complete and clear discussion

## Further comments

One last suggestion on the manuscript would be to rework the structure of headings by removing the numbers in front of the sections. I guess like before, I am just going off of what Dr. Handel said about ready for publication and I feel like most stuff I've worked on hasn't had numbers in front of sections and sub sections, but that could have just been the publications we were going for. Also might want to make your heading a little bit smaller.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

Looks great! Very organized.

### Summary assessment

* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

As I said before, I think you did a great job walking the reader or reviewer through your steps in both the manuscript and the code. I do have a suggestion for the README files though. A few of them were empty and overall, based on Dr. Handel's examples, I think he might want more description here. Also, one of them reads like this...

"Run the analysisscript.rmd to generate an html
Exercise 10 begins on line 366 "

I'm guessing that is left over from the template or an accidental paste. Either way, you probably want to get that removed.

### Summary assessment
* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

Everything ran without any errors!

### Summary assessment
* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments
I feel like you went above and beyond with both the desriptive analysis and the modeling. I would say this is very thorough.

### Summary assessment
* strong level of thorougness

## Further comments

_Add any other comments regarding the overall project here. Write anything you think can help your classmate improve their project._
